---
title: "Time of Day / Delay Group"
author: "Jayne Hoyt"
date: "`r format(Sys.time(), '%B, %d, %Y')`"
output:
  html_document: "default"
  pdf_document: "default"
  theme: "journal"
editor_options: 
  chunk_output_type: inline
---

```{r}
library(ggplot2)
library(dplyr)
library(car) 
library(stats)
library(paletteer)

Air_data <- read.csv("FinalData.csv")


#View(Air_data)
range(Air_data$DepTime)
```

```{r}
#Create Variable for Departure Hours Night vs Day

Air_data$DepTime <-formatC(Air_data$DepTime, width = 4, format = "d", flag = "0")  #Makes sure each entry is length 4, adds 0s if not
Air_data$DepTime <- substr(Air_data$DepTime, start = 1, stop = 2)  #Extracts first two characters to get hours only
Air_data$DepTime <- as.numeric(Air_data$DepTime)
Air_data$DepTime[Air_data$DepTime == 24] <- 00   # Make sure we only have one variable for midnight
  
Air_data <- mutate(Air_data, night = (DepTime < 06 | DepTime > 19)) #Between 7pm and 6am
head(Air_data)
```
# Create Linear Model for `night`

```{r}
air_lm <-  lm(TaxiOut ~ night, data = Air_data)
```

# Wilcox Test

Determine if there is a difference in means between "night" time and "day" time


```{r}
#ANOVA for night vs day

wilcox.test(TaxiOut ~ night, alternative = c("greater"), data = Air_data)
Air_data %>% group_by(night) %>% summarize(mean(TaxiOut))
```

We can see there is a significant difference in means between day time and night time.

```{r}
# Graph for Time of Departure

library(ggrepel)

Air_data_avg_time <- Air_data %>%
  group_by(DepTime) %>%
  summarise(mean_Taxi = mean(TaxiOut, na.rm = TRUE))

timePlot <- ggplot(data=Air_data_avg_time, aes(x=DepTime, y=mean_Taxi))  +
  geom_line()+
  geom_point()+ 
  labs(x = "Time of Day (Hour)", y = "Taxi Out Time (minutes)")+
  theme_classic() +
  scale_fill_manual(values = paletteer_c("grDevices::Purples 3", 30))+
  theme(legend.position = "none", axis.title = element_text(size= 20, face= "bold"), axis.text=element_text(size=18))

timePlot

```



```{r}
# See if time `05` where spike occurs is significantly different from the other times

depTime.aov <- aov(TaxiOut ~ as.factor(DepTime), data = Air_data)
depTimeTukey <- TukeyHSD(depTime.aov)
depTimeTukey
```

We can see a significant difference between time `5` and other times.


#######################

# Departure Delay Groups

```{r}
#Categorize by Early and Late

Air_data <- mutate(Air_data, notDelayed = (DepartureDelayGroups <= 0)) #Less than or equal to 0
#View(Air_data)
```

# Create Linear Model with `early`

```{r}
air_lm <-  lm(TaxiOut ~ notDelayed, data = Air_data)
```

# Wilcox Test

Determine if there is a difference in means between `early` flights and `delayed` flights


```{r}
#ANOVA for Delay Groups "notDelayed" vs "delayed"

wilcox.test(TaxiOut ~ notDelayed, alternative = c("greater"), data = Air_data)
Air_data %>% group_by(notDelayed) %>% summarize(mean(TaxiOut))
```
```{r}
# Graph for `DepartureDelayGroups`

Air_data$DepartureDelayGroups <- as.numeric(Air_data$DepartureDelayGroups)

Air_data_Delay <- Air_data %>%
  group_by(DepartureDelayGroups) %>%
  summarise(mean_Taxi = mean(TaxiOut, na.rm = TRUE))

DelayPlot2 <- ggplot(data=Air_data_Delay, aes(x=as.factor(DepartureDelayGroups), y=mean_Taxi)) +
  geom_col(stat="identity", fill = c("#4A3D8D", "#4A3D8D", "#4A3D8D", "#8c81d1", "#8c81d1", "#8c81d1", "#8c81d1", "#8c81d1", "#8c81d1","#8c81d1", "#8c81d1", "#8c81d1", "#8c81d1", "#8c81d1", "#8c81d1")) + 
  labs(x = "Departure Delay (by 15 minutes)", y = "Taxi Out Time (minutes)") +
  theme_classic()  +
  theme(legend.position = "none")+
  theme(legend.position = "none", axis.title = element_text(size= 20, face= "bold"), axis.text=element_text(size=18))

DelayPlot <- ggplot(data=Air_data_Delay, aes(x=as.factor(DepartureDelayGroups), y=mean_Taxi)) +
  geom_col(stat="identity", fill = "#4A3D8D") + 
  labs(x = "Departure Delay (by 15 minutes)", y = "Taxi Out Time (minutes)") +
  theme_classic()  +
  theme(legend.position = "none")+
  theme(legend.position = "none", axis.title = element_text(size= 20, face= "bold"), axis.text=element_text(size=18))

DelayPlot

```

###########################

# Four Different Times of Day in `TimeOfDay` Variable

```{r}
#Create the `TimeOfDay` Variable

Air_data$TimeOfDay <- ifelse(Air_data$DepTime >= 02 & Air_data$DepTime < 08, "Morning", #Between 2am and 8am
                            ifelse(Air_data$DepTime >= 08 & Air_data$DepTime < 14, "Day",#Between 8am and 2pm
                            ifelse(Air_data$DepTime >= 14 & Air_data$DepTime < 20, "Evening",  #Between 2pm and 8pm
                                   "Night")))#Between 8pm and 2am
table(Air_data$TimeOfDay)
```

```{r}
#See the different means for each time 

Air_data %>% group_by(TimeOfDay) %>% summarize(mean(TaxiOut))
```

```{r}
# ANOVA
time_lm <- lm(TaxiOut ~ TimeOfDay, data = Air_data)
anova(time_lm)

time_aov <- aov(TaxiOut ~ TimeOfDay, data = Air_data)
TukeyHSD(time_aov)
```
We can see here that:

Morning = highest
Day 
Evening
Night = lowest